The role of advection versus diffusion in intraventricular transport

### 2025-02-04 paper/main.tex
Results 5 and 6

\subsection*{Numerical approximation of the advection-diffusion equation}
The advection-diffusion equation~\eqref{eq:adv_diff_strong} is discretized
with a first-order implicit Euler method in time, and a first-order,
symmetric interior penalty discontinuous Galerkin (DG)
method in space~\cite{Arnold1982AnElements}.
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.02252$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

@@

\subsection*{Loss of cilia motility in specific subpopulation slows down transport}
We previously identified three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the role of these lineages on CSF transport, we modified flow boundary conditions
in the three different ciliated regions
(\Cref{fig:figure5_compare_cilia_modifications}a, magenta markers)
and simulated CSF flow using the Stokes equations.
We identified that the velocities were most strongly
affected when removing cilia in the dorsal diencephalic cilia,
as the maximum velocity dropped from 27.7 $\mu$m/s to 8.0 $\mu$m/s.
In contrast, removing telencephalic and ventral diencephalic had minimal
impact and resulted in maximum velocities of 27.7 $\mu$m/s or 27.2 $\mu$m/s, respectively.

With the resulting velocity fields, we simulated transport of Dendra2.
Changes in the transport dynamics were most prominent when removing the dorsal cilia
in the diencephalic ventricle, which produces the strongest flow at this developmental stage.
All ROIs (except ROI 1, where we locally released solutes) exhibit significant changes
in the dynamics of $\cbar$ (\Cref{fig:figure5_compare_cilia_modifications}b).
Compared to the baseline model results, we observed a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and a faster transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia did not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:figure5_compare_cilia_modifications}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute has spread towards the anterior ventricle (ROI 5) and
less solute has spread towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:figure5_compare_cilia_modifications}
\end{figure}

\subsection*{Modification of ventricles geometry impacts protein distribution}
To assess the impact of ventricular morphologies on solute distribution,
we used four alternative brain ventricles meshes:
one where the connection between the anterior and middle ventricles is constricted;
one where the connection between the middle and posterior ventricles is constricted;
one where the middle ventricle is shrunk in the lateral direction;
and the combination of the above three modifications (\Cref{fig:figure6_sim_results_compare_geometries}a).
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2 (\Cref{fig:figure6_sim_results_compare_geometries}b).

>>>>

\subsection*{Numerical approximation of the advection-diffusion equation}
The advection-diffusion equation~\eqref{eq:adv_diff_strong} is discretized
with a first-order implicit Euler method in time, and a first-order,
symmetric interior penalty discontinuous Galerkin (DG)
method in space~\cite{Arnold1982AnElements}, chosen due to its favorable
conservation properties.
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.02252$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

@@

\subsection*{Loss of cilia motility in specific subpopulation slows down transport}
We previously identified three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the role of these lineages on CSF transport, we simulate transport
upon region-specific cilia paralysis by modifying the boundary conditions
of the CSF flow model. Three separate cilia-covered regions were considered separately:
the telencephalic (anterior) ventricle, the dorsal part and the ventral part
of the diencephalic (middle) ventricle (\Cref{fig:figure5_compare_cilia_modifications}a).
The cilia tangential traction boundary condition is for each case
changed to a free-slip boundary condition in paralyzed region,
prescribing zero tangential forces to
represent absence of cilia motion.
We simulate CSF flow modeled by the Stokes equations with the
modified boundary conditions, and use the resulting
velocity fields in transport simulations.

We identified that the velocities were most strongly
affected when removing cilia in the dorsal diencephalic cilia,
as the maximum velocity dropped from 27.7 $\mu$m/s to 8.0 $\mu$m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic had minimal
impact and resulted in maximum velocities of 27.7 $\mu$m/s or 27.2 $\mu$m/s, respectively.
Changes in the transport dynamics of Dendra2 were most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit significant changes in the dynamics of $\cbar$
(\Cref{fig:figure5_compare_cilia_modifications}b).
Compared to the baseline model results, we observed a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia did not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:figure5_compare_cilia_modifications}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute has spread towards the anterior ventricle (ROI 5) and
less solute has spread towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:figure5_compare_cilia_modifications}
\end{figure}

\subsection*{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impacts the CSF flow patterns.
To assess the impact of ventricular morphologies on solute distribution,
we used four alternative brain ventricles meshes:
one where the connection between the anterior and middle ventricles is constricted;
one where the connection between the middle and posterior ventricles is constricted;
one where the middle ventricle is shrunk in the lateral direction;
and the combination of the above three modifications (\Cref{fig:figure6_sim_results_compare_geometries}a).
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Cref{fig:figure6_sim_results_compare_geometries}b).
